---
title: "About Class"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(kableExtra)
```

**Office Location**: CBLS 377

**Telephone**: 4-5404

**Email**: rsschwartz@uri.edu  
Please note that I will do my best to respond promptly between 9-5 on weekdays.

**Office Hours**: 1 hour / week TBD  
Please feel free to come by without an appointment during this time. To talk with me during other times please set up a meeting by email.

**Class Days / Time**: M 3-4pm Coastal Institute (Kingston) 117A

**Prerequisites**: BIO 262 (previous or concurrent)

**Required text**:
Whitlock and Schluter. 2015. The analysis of biological data. 2nd ed.

**Other equipment / material requirements**:
Laptop, Tablet, or Chromebook with internet access. Please contact the instructor if you do not have a laptop and purchasing one would be a financial difficulty. Laptops are available for use in class upon request.

**Assignments**:
Each of the four sections will be worth 250 points. For each section, 150 points will be earned based on follow-ups each week to in-class work. 100 points will be earned by consolidating the results into a short report. Assignments will build on skills and concepts learned in previous weeks. Assignments will be initiated in class and are due before class the following week.

This course will require engagement with material, critical thinking, and the ability to communicate. The Academic Enhancement Center helps URI students succeed through three services: Academic Coaching, Subject-Based Tutoring, and The Writing Center. To learn more about any of these services, please visit http://uri.edu/aec or call 401-874-2367 to speak with reception staff. Hours are posted at http://web.uri.edu/ceps/academic-skills-center/

**Grades**:

```{r}
Grade <- c("A", "A-", "B+", "B", "B-", "C+", "C", "C-", "D+", "D", "F") 
Points <- c("940-1000", "900-939", "870-899", "830-869", "800-829", "770-799", "730-769", 
            "700-729", "670-699", "600-669", "<600")

kable(cbind(Grade,Points), "html") %>%  
  kable_styling(bootstrap_options = c("striped", "hover", "responsive"))

```

**Special note**:
If you, or anyone you know, is at risk of going hungry please visit the Rhody Outpost, a food pantry operated by URI. Rhody Outpost is open twice per month at St. Augustine Church at 15 Lower College Rd. For open hours see https://web.uri.edu/deanofstudents/rhody-outpost/ . Rhody Outpost protects the privacy of all visitors. Students may also visit the Jonnycake Center in Peace Dale.